---
title: "10xVisium_BreastCancer"
author: "Lauren Rusnak"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prep

```{r include=F}
library(SpatialExperiment)
library(STexampleData)
library(scran)
library(DropletUtils)
library(ggplot2)
library(ggspavis)
library(scater)
library(smoothclust)
library(stats)
library(mclust)
library(randnet)
library(R.utils)
library(tibble)
library(dplyr)
```



## Import Data

_________________________________________________________________________



```{r}
dir <- file.path("C:/Users/lerus/OneDrive/Documents/BU-AB/ResearchRotation/10xVisium")
  
sample_ids <- c("Visium_Human_Breast_Cancer")
samples <- file.path(dir, sample_ids)

spe <- read10xVisium(samples)

# tabulate number of spots mapped to tissue
cd <- colData(spe)
table(
  in_tissue = cd$in_tissue, 
  sample_id = cd$sample_id)

# view available images
imgData(spe)
```

```{r}
spe
```


_______________________________________________________________________________


### Normalization and Feature Selection

*Normalization*
```{r}
spe <- logNormCounts(spe) 
assayNames(spe)
```

*Calculate highly variable genes (HVGs) and select the top HVGs*
```{r}
dec <- modelGeneVar(spe) 
top_hvgs <- getTopHVGs(dec, prop = 0.1)
spe <- spe[top_hvgs, ]
dim(spe)
```



_________________________________________________________________________


### **Run smoothclust with different parameters**

#### Method: Uniform

```{r, include = FALSE}
# library(smoothclust)
spe <- smoothclust(spe, method = "uniform")
```

```{r}
assayNames(spe)
```



##### Dimensionality reduction

```{r}
# Principal Component Analysis
set.seed(123)
spe <- runPCA(spe, subset_row = top_hvgs, exprs_values = "logcounts_smooth")

# UMAP on top 50 PCs -> for visualization
set.seed(123)
spe <-  runUMAP(spe, dimred= "PCA", exprs_values = "logcounts_smooth")

dim(reducedDim(spe, "PCA"))
dim(reducedDim(spe, "UMAP"))

# update column names
colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)

# Vector of principal commponents for clustering 
pcs <- reducedDim(spe, "PCA")
```


```{r}
n_clusters <- 15
# Initialize total within sum of squares error: wss
wss <- numeric(n_clusters)
set.seed(123)

# Look over 1 to n possible clusters
for (i in 1:n_clusters) {
  # Fit the model: km.out
  km.out <- kmeans(pcs, centers = i, nstart = 20)
  # Save the within cluster sum of squares
  wss[i] <- km.out$tot.withinss
}

# Produce a scree plot
wss_df <- tibble(clusters = 1:n_clusters, wss = wss)
 
scree_plot <- ggplot(wss_df, aes(x = clusters, y = wss, group = 1)) +
    geom_point(size = 4)+
    geom_line() +
    xlab('Number of clusters') +
    scale_x_continuous(breaks = c(2,4,6,8,10,12,14,16))
scree_plot
```


##### K-means clustering with 4 centers
```{r}
## 4 centers

set.seed(123)

km.4 <- kmeans(pcs, centers = 4)

table(km.4$cluster)

# kmeans cluster assignments
km.clus4 <- km.4$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus4)
```


###### **Visualization**

```{r}
# plot clusters in spatial x-y coordinates
xyplot_uni_km3 <- plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.9)
xyplot_uni_km3
```


*Reduced Dimensionality Cluster Visualization*
```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)

# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```



##### K-means clustering with 7 centers
```{r}
## 7 centers (one fore each true layer + 1 for white matter)

set.seed(123)

km.7 <- kmeans(pcs, centers = 7)

table(km.7$cluster)

# kmeans cluster assignments
km.clus7 <- km.7$cluster

# store cluster label in column 'label' in colData
colLabels(spe) <- factor(km.clus7)
```


###### **Visualization**

```{r}
# plot clusters in spatial x-y coordinates
xyplot_uni_km7 <- plotSpots(spe, annotate = "label", 
          palette = "libd_layer_colors",
          size = 0.7)
xyplot_uni_km7
```


*Reduced Dimensionality Cluster Visualization*
```{r}
# plot clusters in PCA reduced dimensions
plotDimRed(spe, type = "PCA", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)

# plot clusters in UMAP reduced dimensions
plotDimRed(spe, type = "UMAP", 
           annotate = "label", palette = "libd_layer_colors",
           size = 0.7)
```


